# Interactive Graphing: Discharge of the Poudre River 

>"Someone asked me to name two structures that hold water. I was like, 'well… damn!'" 

This assignment used a unique package of R Markdown (dygraphs) in order to create an interactive chart.

Data and assignment provided by Dr. Matthew Ross and Dr. Nathan Mueller of Colorado State University.

```{r Ch2 setup, include=FALSE}
library(tidyverse)
library(dataRetrieval)
library(dygraphs) #this is for interactive
library(xts)
```

## Background on the Poudre River

[Cache La Poudre River](https://edits.nationalmap.gov/apps/gaz-domestic/public/summary/205018) is an important watershed that supports **agriculture, industry, recreation, and residential needs** on the Front Range of Colorado. It also provides for cottonwood forest, shrub, and grassland ecosystems that support [wildlife](http://poudretrail.org/habitat-wildlife/#fish) from the mountains down to the prairies. The unique **biodiversity** and **history** of the Cache La Poudre watershed are valued widely; 45 miles along the Poudre are encompassed in a [National Heritage Area](https://www.nps.gov/places/cache-la-poudre-river-national-heritage-area.htm).
The history of Cache La Poudre is linked to the *history of the West*, because its banks supported the first major [irrigation-based agricultural settlement](https://www-jstor-org.ezproxy2.library.colostate.edu/stable/pdf/1821074.pdf) of its kind in 1870, which would soon spread through the Arid West.

## Interactive Discharge Chart

```{r interactive, fig.align='center', fig.cap="Discharge of the Poudre River in cubic feet per second from January 2017 to December 2021."}

q <- readNWISdv(siteNumbers = '06752260',
                parameterCd = '00060',
                startDate = '2017-01-01',
                endDate = '2022-01-01') %>%
  rename(q = 'X_00060_00003')

q_xts <- xts(q$q, order.by = q$Date)

dygraph(q_xts) %>%
  dyAxis("y", label = "Discharge (cfs)") %>%
  dyOptions(drawPoints = TRUE, pointSize = 2)

```
