# Extracting and Visualizing Meteorological Data

>"What do you call dangerous precipitation? A rain of terror."

For this assignment, we used custom functions to read in and look at average meteorological data scraped from a public data archive.

Data is from [Snowstudies.org](https://snowstudies.org/archived-data/). Assignment by Dr. Matthew Ross and Dr. Nathan Mueller of Colorado State University.

```{r Ch5 setup, include=FALSE}
library(rvest)
library(tidyverse)
library(lubridate)
library(readxl)

```

## Extract the meteorological data URLs. 

```{r, warning=FALSE}

# Read HTML page 
snowarchive <- read_html("https://snowstudies.org/archived-data/")

# Read link with specific pattern
links <- snowarchive %>%
  html_nodes('a') %>% #look for links
  .[grepl('forcing',.)] %>% #filter to only links with "forcing" term
  html_attr('href') #tell it these are urls

links # view

```

## Download the meteorological data from the URL

```{r, warning=FALSE}

# Grab only the name of the file by splitting out on forward slashes
splits <- str_split_fixed(links,'/',8)

#Keep only the 8th column
files <- splits[,8] 

files

# Generate a file list for where the data goes
file_names <- paste0('Data_sci_bookdown/data/snow/', files)

# For loop that downloads each - i for every instance, length function tells how many instances
for(i in 1:length(file_names)){
  download.file(links[i],destfile=file_names[i])
}

# Download via map function
#map2(links, file_names, download.file)

# Map version of the for loop (downloading files)
downloaded <- file.exists(file_names) 
evaluate <- !all(downloaded) # sees if files are downloaded (T/F)
if(evaluate == T){
  map2(links[1:2],file_names[1:2],download.file)
}else{print('data downloaded')}

```

## Write a custom function to read in the data and append a site column to the data

```{r, warning=FALSE}

# Traditional read in

SASP <- read.csv("Data_sci_bookdown/data/snow/SBB_SASP_Forcing_Data.csv") %>%
  select(1,2,3,7,10)
  
colnames(SASP) <- c("year","month","day","precip","temp")

SBSP <- read.csv("Data_sci_bookdown/data/snow/SBB_SBSP_Forcing_Data.csv") %>%
  select(1,2,3,7,10)
  
colnames(SBSP) <- c("year","month","day","precip","temp")

# Combine csvs
alldata <- rbind(SASP,SBSP)

# Read in via new function

# Grab headers from metadata pdf
library(pdftools)
headers <- pdf_text('https://snowstudies.org/wp-content/uploads/2022/02/Serially-Complete-Metadata-text08.pdf') %>%
  readr::read_lines(.) %>%
  trimws(.) %>%
  str_split_fixed(.,'\\.',2) %>%
  .[,2] %>%
  .[1:26] %>%
  str_trim(side = "left")

```

## Use the `map` function to read in both meteorological files

```{r, warning=FALSE}

# Pull site name out of the file name and read in the .txt files
read_data <- function(file){
  name = str_split_fixed(file,'_',2)[,2] %>% 
    gsub('_Forcing_Data.txt','',.) 
  df <- read_fwf(file) %>% 
    select(year=1, month=2, day=3, hour=4, precip=7, air_temp=10) %>% #choose and name columns
    mutate(site = name) #add column 
}

alldata2 <- map_dfr(file_names,read_data) 

summary(alldata2)

```

## Make a line plot of mean temp by year by site

```{r, warning=FALSE, fig.cap="Mean temperature of the SASP (teal) and SBSP (orange) sites from 2003 to 2012, in degrees Kelvin."}

temp_yearly <- alldata2 %>% 
group_by(year, site) %>%
summarise(mean_temp = mean(`air_temp`, na.rm=T))

ggplot(temp_yearly,aes(x=year, y=mean_temp, color=site)) + 
  geom_point() + geom_line() +
  xlab("Year") + ylab("Mean Temperature (Degrees Kelvin)") +
  ggthemes::theme_few() + 
  scale_color_brewer(palette = "Set2") + 
  scale_x_continuous(breaks = pretty(c(2003,2012), n = 6)) +
  theme(legend.position="bottom")

```

## Write a function that makes line plots of monthly average temperature at each site for a given year. Use a for loop to make these plots for 2005 to 2010. 

```{r, warning=FALSE, fig.cap="Mean monthly temperatures in degrees Kelvin for SASP (teal) and SBSP (orange) sites in 2005, 2006, 2007, 2008, 2009, and 2010."}

temp_monthly <- alldata2 %>%
    group_by(year, month, site) %>%
    summarize(mean_temp = mean(`air_temp`, na.rm=T))

par(mfrow=c(5,1))

plot_monthly <- function(year.no) {
  plot <- temp_monthly %>%
    filter(year == year.no) %>%
    ggplot(aes(x=month, y=mean_temp, color=site)) + 
      geom_line() +
      xlab("Month") + ylab("Mean Temperature (Degrees Kelvin)") +
      ggthemes::theme_few() + 
      scale_color_brewer(palette = "Set2") + 
      scale_x_discrete(limits = c(1,2,3,4,5,6,7,8,9,10,11,12)) +
      scale_y_continuous(breaks = pretty(c(255,290), n = 4)) +
      theme(legend.position="bottom")
  print(plot)
  }

for(i in 2005:2010){
  plot_monthly(i)
}

```

## Make a plot of average daily precipitation by day of year (averaged across all available years)

```{r, warning=FALSE, fig.cap="Mean daily precipitation by day of year, averaged from 2003 to 2012."}

precip_daily <- alldata2 %>%
  mutate(date = make_date(year, month, day),
                day_no = yday(date)) %>%
  group_by(day_no) %>%
  summarize(mean_precip = mean(`precip`*86400, na.rm=T))

ggplot(precip_daily, aes(x=day_no, y=mean_precip)) + 
      geom_line() +
      xlab("Day of Year") + ylab("Mean Precipitation (mm/day)") +
      ggthemes::theme_few() + 
      scale_color_brewer(palette = "Set2") + 
      scale_y_continuous(breaks = pretty(c(0,14), n = 7)) +
      scale_x_continuous(breaks = pretty(c(1,365), n = 8))

```