# Spatial Analysis in R

>"Why are latitude and longitude so smart? Because they have so many degrees!"

In this assignment, I learned to use R for spatial analyses.

Data is from the [LAGOS dataset](https://lagoslakes.org/). Assignment by Dr. Matthew Ross and Dr. Nathan Mueller of Colorado State University.

```{r Ch6 setup, warning=FALSE, include=FALSE}
library(tidyverse) # Tidy packages
library(dplyr)
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(LAGOSNE) #Lots and lots of clean lake data

#install.packages("remotes") #installs package to install via github
#remotes::install_github("ropensci/USAboundariesData") #installs USAboundaries

library(USAboundaries) #USA states and counties

```

## Loading in data

### First download and then specifically grab the locus (or site lat longs)

```{r data-read}
# #Lagos download script
#LAGOSNE::lagosne_get(dest_folder = LAGOSNE:::lagos_path(), overwrite = TRUE)


#Load in lagos
lagos <- lagosne_load()

#Grab the lake centroid info
lake_centers <- lagos$locus

# Make an sf object 
spatial_lakes <- st_as_sf(lake_centers,coords=c('nhd_long','nhd_lat'),
                          crs=4326)

#Grab the water quality data
nutr <- lagos$epi_nutr

#Look at column names
#names(nutr)

```

### Convert to spatial data
```{r}
#Look at the column names
#names(lake_centers)

#Look at the structure
#str(lake_centers)

#View the full dataset
#View(lake_centers %>% slice(1:100))

spatial_lakes <- st_as_sf(x = lake_centers, coords = c("nhd_long","nhd_lat"), crs = 4326) %>%
  st_transform(2163)

#mapview(spatial_lakes)

#Subset for plotting
subset_spatial <- spatial_lakes %>%
  slice(1:100) 

subset_baser <- spatial_lakes[1:100,]

#Dynamic mapviewer
#mapview(subset_spatial)

```

### Subset to only Minnesota

```{r}
states <- us_states()

#Plot all the states to check if they loaded
#mapview(states)

minnesota <- states %>%
  filter(name == 'Minnesota') %>%
  st_transform(2163)
#mapview(minnesota)

#Subset lakes based on spatial position
minnesota_lakes <- spatial_lakes[minnesota,]

#Plotting the first 1000 lakes
minnesota_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000)
  #mapview(.,zcol = 'lake_area_ha')
```

## Part one

### Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)

```{r}
Istates <- states %>%
  filter(name == 'Iowa'| name== 'Illinois') %>%
  st_transform(2163)
mapview(Istates, canvas = TRUE) 
```

### Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?

```{r}

Istates_lakes <- spatial_lakes[Istates,]

nrow(Istates_lakes)
Istates_count <- length(Istates_lakes$lagoslakeid)

nrow(minnesota_lakes)
Minn_count <- length(minnesota_lakes$lagoslakeid)

```

Iowa and Illinois have `r Istates_count` lakes combined, much less than the number of lakes that Minnesota alone has, `r Minn_count`.


### What is the distribution of lake size in Iowa vs. Minnesota?

- Here I want to see a histogram plot with lake size on x-axis and frequency on y axis (check out geom_histogram)

```{r, fig.cap="The number of lakes with a given area, in hectares, in Minnesota (red) and Iowa (blue)."}

iowa <- states %>%
  filter(name == 'Iowa') %>%
  st_transform(2163)

iowa_lakes <- spatial_lakes[iowa,]

combined <- rbind(iowa_lakes, minnesota_lakes)

ggplot(combined, aes(x= lake_area_ha)) + 
  ggthemes::theme_few() + theme(legend.position="bottom") +
  xlab("Lake Area (ha)") + ylab("Count") +
  scale_x_continuous(trans = "log10", labels = scales::comma) +
  geom_histogram(data = minnesota_lakes, color = "red", alpha = 0.2) + 
  geom_histogram(data = iowa_lakes, color = "blue", alpha = 0.2) +
  scale_fill_manual(values=c("blue","red"), "State")

```

### Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares

```{r, fig.cap="Lakes in Iowa and Illinois, colored by area in hectares."}
Istates_map = Istates_lakes %>%
  arrange(-lake_area_ha) %>%
    slice(1:1000)

mapview(Istates_map, zcol = 'lake_area_ha',  canvas = TRUE) 

```


### What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 

We might use the US Geological Survey (USGS) National Water Informational System (NWIS) and its National Water Dashboard as a data source, and look at gage height (indicating lake depth) as another parameter for lake size variation. The USGS National Hydrography Dataset (NHD) is another data source that would, similarly to Lagos, give us a surface area metric for lakes in the various states.

## Part two

### Subsets 

#### Columns nutr to only keep key info that we want

```{r}
clarity_only <- nutr %>%
  dplyr::select(lagoslakeid,sampledate,chla,doc,secchi) %>%
  mutate(sampledate = as.character(sampledate) %>% ymd(.))

```

#### Keep sites with at least 200 observations 

```{r}

#Look at the number of rows of dataset
#nrow(clarity_only)

chla_secchi <- clarity_only %>%
  filter(!is.na(chla),
         !is.na(secchi))

# How many observatiosn did we lose?
# nrow(clarity_only) - nrow(chla_secchi)


# Keep only the lakes with at least 200 observations of secchi and chla
chla_secchi_200 <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n()) %>%
  filter(count > 200)


```

#### Join water quality data to spatial data

```{r}
spatial_200 <- inner_join(spatial_lakes,chla_secchi_200 %>%
                            distinct(lagoslakeid,.keep_all=T),
                          by='lagoslakeid')

```

### Mean Chlorophyll A map

```{r, fig.cap="Mean chlorophyl content at sites with at least 200 observations."}
### Take the mean chl_a and secchi by lake

means_200 <- chla_secchi_200 %>%
  # Take summary by lake id
  group_by(lagoslakeid) %>%
  # take mean chl_a per lake id
  summarize(mean_chl = mean(chla,na.rm=T),
            mean_secchi=mean(secchi,na.rm=T)) %>%
  #Get rid of NAs
  filter(!is.na(mean_chl),
         !is.na(mean_secchi)) %>%
  # Take the log base 10 of the mean_chl
  mutate(log10_mean_chl = log10(mean_chl))

#Join datasets
mean_spatial <- inner_join(spatial_lakes,means_200,
                          by='lagoslakeid') 

#Make a map
mapview(mean_spatial, zcol='log10_mean_chl', layer.name = "Mean Chlorophyll A Content")
```

### What is the correlation between Secchi Disk Depth and Chlorophyll a for sites with at least 200 observations?

```{r, fig.cap="Chlorophyll content has a negative correlation with Secchi disk depth at sites with at least 200 observations."}
ggplot(means_200) +
  geom_point(aes(mean_secchi, mean_chl)) +
  ggthemes::theme_few() + 
  xlab("Mean Secchi Disk Depth") + ylab("Mean Chlorophyll Content")

```

#### Why might this be the case? 

Secchi disks measure water clarity; the deeper the disk, the clearer the water (1). Chlorophyll content in lakes is generally a reliable marker of algae content, so that high chlorophyll values indicate high algal biomass and corresponding low water clarity (2). Additionally, chlorophyll may be used as a proxy for water quality, since high algal biomass is associated with high nutrient pollution in the process of eutrophication (2). High pollution may further decrease water clarity, so that the relationship between chlorophyll and Secchi disk depth may be expected.

1. “The Secchi Dip-in - What Is a Secchi Disk?” North American Lake Management Society (NALMS),
https://www.nalms.org/secchidipin/monitoring-methods/the-secchi-disk/what-is-a-secchi-disk/. 

2. Filazzola, A., Mahdiyan, O., Shuvo, A. et al. A database of chlorophyll and water chemistry in freshwater lakes. Sci Data 7, 310 (2020).
https://doi-org.ezproxy2.library.colostate.edu/10.1038/s41597-020-00648-2


### What states have the most data? 

#### Make a lagos spatial dataset that has the total number of counts per site.

```{r}

site_counts <- chla_secchi %>%
  group_by(lagoslakeid) %>%
  mutate(count = n())

lake_counts <- inner_join(site_counts, lake_centers, by= "lagoslakeid")%>% 
  dplyr::select(lagoslakeid,nhd_long,nhd_lat, count, secchi, chla)

spatial_counts <- st_as_sf(lake_counts,coords=c("nhd_long","nhd_lat"),
                          crs=4326)

```


#### Join this point dataset to the us_boundaries data. 

```{r}

states <- us_states()

states_counts <- st_join(spatial_counts, states)

```


#### Group by state and sum all the observations in that state and arrange that data from most to least total observations per state. 

```{r}

sum_statecount <- states_counts %>%
  group_by(state_name) %>%
  summarize(sum = sum(count)) %>%
  arrange(desc(sum)) 

sumtable <- tibble(sum_statecount)

view(sumtable)

#ggplot(data = sumtable, aes(x=state_name, y=sum, fill=state_name)) + 
#  geom_bar(stat = "identity", width = 0.3, position = "dodge") + 
#  ggthemes::theme_few() +
#  xlab("State") + ylab(expression(paste("# of Observations"))) 


```

Minnesota has the most observations. Vermont, has the next most observations, but less than half of Minnesota's observations. South Dakota has the least number of observations in the dataset.

### Is there a spatial pattern in Secchi disk depth for lakes with at least 200 observations?

```{r, fig.cap="Map showing mean Secchi disk depths of sites with at least 200 observations."}
mapview(mean_spatial, zcol='mean_secchi', layer.name = "Mean Secchi Disk Depth")

```

