# Evaluating Model Predictions

>"Being a statistician means never having to say you are certain."



Data and assignment provided by Dr. Michael Lefsky of Colorado State University.

```{r Ch9 setup, warning=FALSE, echo=FALSE, message=FALSE}

knitr::opts_knit$set(root.dir='..')

library(tidyr)
library(readr) #read files
library(dplyr) #lots of functions
library(ggplot2) #all of the graphs
library(ggthemes) #themes for ggplots
library(lme4) #mixed models
library(lmerTest)
library(emmeans)
library(plotrix)
library(car) #ANOVA
library(MASS)
library(RColorBrewer) #color pallets for graphs
library(Metrics) #does rmse function
library(mgcv) #can do generalized additive models
library(MuMIn) #Multi Model Inference - AIC, GME

```

## Merge model predictions and observed data

```{r, warning=FALSE, message=FALSE}

# Load datasets

#Century Model output through 2016
centgrain <- read.csv("Data_sci_bookdown/data/model-assess/century_harvest.csv")

centgrain13 <- centgrain[-c(16:18),] #Century model outputs through 2013

centwheat <- centgrain13 %>%
  filter(Crop == "wheat") #just the wheat outputs

centcorn <- centgrain13 %>%
  filter(Crop == "corn") #just the corn outputs

#load observed data
obsgrain <- read.csv("Data_sci_bookdown/data/model-assess/obs_corn_wheat_cgrain.csv") #observed data through 2013

obswheat <- obsgrain %>%
  filter(Crop == "wheat") #just the wheat outputs

obscorn <- obsgrain %>%
  filter(Crop == "corn") #just the corn outputs

#merge data from Century Model and observations
wheat <- merge(centwheat,obswheat, by=c('Year','Crop'), all.x = T) # merge predicted and observed wheat yields
corn <- merge(centcorn,obscorn, by=c('Year','Crop'), all.x = T) # merge predicted and observed wheat yields
allgrain <- merge(centgrain13, obsgrain, by=c('Year','Crop'), all.x = T) # merge all predicted and observed

```

## Linear regression model and ANOVA of wheat predictions vs. observations

```{r, warning=FALSE, fig.cap="Scatterplot and linear regression line of predicted and observed wheat yields in grams of carbon (C) per m2 per year between 1999 and 2013. The equation of the line is -0.436x + 512, demonstrating that observed yields were lower than predicted yields on average. The multiple R2 value is 0. 201 and the adjusted R2 value is 0.0683, indicating a very weak linear relationship. The p-value is 0.265, considerably higher than what is generally considered to be significant."}

lm_wheat <- lm(cgrain_cent ~ cgrain_obs, data = wheat) # linear model of predicted vs. observed wheat yields
summary(lm_wheat) # intercept, slope (estimate column), P, R^2
summary(aov(lm_wheat)) # ANOVA

ggplot(data = wheat, aes(x=cgrain_cent,y=cgrain_obs)) + 
  geom_point(color="black") + geom_smooth(method="lm", se=FALSE, color="#78917E") +
  xlab("Predicted Yields (g C/m2/yr)") + ylab(expression(paste("Observed Yields (g C/m2/yr)"))) + 
  ggtitle("Predicted vs. Observed Wheat Yields") +
  theme_few(base_size = 16)

```

## Linear regression model and ANOVA of corn predictions vs. observations

```{r, warning=FALSE, fig.cap="Scatterplot and linear regression line of predicted and observed corn yields in grams of carbon (C) per m2 per year between 1999 and 2013. The equation of the line is 0.320x + 282, demonstrating that observed yields were higher than predicted yields on average. The multiple R2 value is 0.406 and the adjusted R2 value is 0.287, indicating that the linear relationship is weak. The p-value is 0.124, a little higher than what is generally considered to be significant."}

lm_corn <- lm(cgrain_cent ~ cgrain_obs, data = corn) # linear model of predicted vs. observed corn yields
summary(lm_corn) # intercept, slope (estimate column), P, R^2
summary(aov(lm_corn)) # ANOVA

ggplot(data = corn, aes(x=cgrain_cent,y=cgrain_obs)) + 
  geom_point(color="black") + geom_smooth(method="lm", se=FALSE, color="#78917E") +
  xlab("Predicted Yields (g C/m2/yr)") + ylab(expression(paste("Observed Yields (g C/m2/yr)"))) + 
  ggtitle("Predicted vs. Observed Corn Yields") +
  theme_few(base_size = 16)

```

## Mean, standard deviation, and standard error for predicted and observed outputs

```{r, warning=FALSE}

#Summary stats for predicted outputs - mean, sd, se
centwheat %>% # summary stats for CM wheat outputs
  summarise(n = n(),
            mean = mean(cgrain_cent), # mean
            sd = sd(cgrain_cent), # standard deviation
            SE = sd/sqrt(n)) # standard error

centcorn %>% # summary stats for CM corn outputs
  summarise(n = n(),
            mean = mean(cgrain_cent), # mean
            sd = sd(cgrain_cent), # standard deviation
            SE = sd/sqrt(n)) # standard error

#Summary stats for observed outputs - mean, sd, se
obswheat %>% # summary stats for observed wheat outputs
  summarise(n = n(),
            mean = mean(cgrain_obs), # mean
            sd = sd(cgrain_obs), # standard deviation
            SE = sd/sqrt(n)) # standard error

obscorn %>% # summary stats for observed corn outputs
  summarise(n = n(),
            mean = mean(cgrain_obs), # mean
            sd = sd(cgrain_obs), # standard deviation
            SE = sd/sqrt(n)) # standard error
```

## Model evaluation via Mean Absolute Error (MAE), Root Mean Squared Error (RMSE), and Model Efficiency (ME)

```{r, warning=FALSE}

#Model evaluation stats
mae(centgrain13[,3], obsgrain[,3]) # mean absolute error for predicted vs. observed yields
rmse(centgrain13[,3], obsgrain[,3]) # root mean squared error for predicted vs. observed yields

```


