--- 
title: "We R Under Way: A Data Science Portfolio"
author: "Kaydee Barker"
date: "03/21/2022"
site: bookdown::bookdown_site
output: bookdown::gitbook
always_allow_html: yes
documentclass: book
bibliography: [packages.bib]
biblio-style: apalike
link-citations: yes
description: "Just some of my adventures so far in the lands of environmental data. Warning: I can't help adding nerd jokes, so pun-ish me."
---
--- 
title: "We R Under Way: A Data Science Portfolio"
author: "Kaydee Barker"
date: "03/21/2022"
site: bookdown::bookdown_site
output: bookdown::gitbook
always_allow_html: yes
documentclass: book
bibliography: [packages.bib]
biblio-style: apalike
link-citations: yes
description: "Just some of my adventures so far in the lands of environmental data. Warning: I can't help adding nerd jokes, so pun-ish me."
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# automatically create a bib database for R packages
knitr::write_bib(c(
  .packages(), 'bookdown', 'knitr', 'rmarkdown'
), 'packages.bib')

```

Writing and code by Kaydee Barker, assignments by Dr. Ross and Dr. Mueller (SOCR 580A7), Dr. Lefsky (ESS 330) of Colorado State University. Data cited within chapters.



<!--chapter:end:index.Rmd-->

# Introduction {#intro}

>"There are two kinds of data scientists: 1) Those who can extrapolate from incomplete data." 


I began my foray into R in the spring of 2020, first teaching myself some basic syntax and then using it for statistical analysis on my research projects as an undergraduate researcher at Colorado State University (CSU). With the help of my research mentors and many amazing people of the internet, I was able to fumble my way forward and learn a number of techniques to analyze and visualize data in R. I have since been building on my R and data science skills, including with the help of two key courses at CSU: "Quantitative Reasoning for Ecosystem Science" (ESS 330) and "Introduction to Environmental Data Science" (SOCR 580A7). Since I can't yet publish data from my research projects, this portfolio is constructed of public data examples, primarily from my coursework in those two courses. Its purpose is a) to serve as a reference for myself and others learning to use R for environmental analyses, and b) to demonstrate my current R knowledge to advisors and colleagues.

<!--chapter:end:01-Introduction.Rmd-->


# Interactive Graphing: Discharge of the Poudre River 

Placeholder


## Background on the Poudre River
## Interactive Discharge Chart

<!--chapter:end:02-Poudre-Interactive.Rmd-->


# Looking at Effects of Fire on Vegetation

Placeholder


## Introduction
## What is the correlation between NDVI and NDMI?
## What month is the greenest month on average? 
## What month is the snowiest on average?

<!--chapter:end:03-Correlations-Fire.Rmd-->


# Fire Effects on Fish Populations

Placeholder


## Pre versus post fire fish length and mass
## Linear regression of fish mass vs. length for before and after the fire

<!--chapter:end:04-Fire-Fish-Populations.Rmd-->


# Extracting and Visualizing Meteorological Data

Placeholder


## 1. Extract the meteorological data URLs. Here we want you to use the `rvest` package to get the URLs for the `SASP forcing` and `SBSP_forcing` meteorological datasets.
## 2. Download the meteorological data. Use the `download_file` and `str_split_fixed` commands to download the data and save it in your data folder. You can use a for loop or a map function. 
## 3. Write a custom function to read in the data and append a site column to the data. 
## 4. Use the `map` function to read in both meteorological files. Display a summary of your tibble.
## 5. Make a line plot of mean temp by year by site (using the `air temp [K]` variable). Is there anything suspicious in the plot? Adjust your filtering if needed.
## 6. Write a function that makes line plots of monthly average temperature at each site for a given year. Use a for loop to make these plots for 2005 to 2010. 
## Bonus: Make a plot of average daily precipitation by day of year (averaged across all available years)

<!--chapter:end:05-Extract-Data.Rmd-->


# Spatial Analysis in R

Placeholder


## Loading in data
### First download and then specifically grab the locus (or site lat longs)
### Convert to spatial data
### Subset to only Minnesota
## Part one
### Show a map outline of Iowa and Illinois (similar to Minnesota map upstream)
### Subset LAGOS data to these sites, how many sites are in Illinois and Iowa combined? How does this compare to Minnesota?
### What is the distribution of lake size in Iowa vs. Minnesota?
### Make an interactive plot of lakes in Iowa and Illinois and color them by lake area in hectares
### What other data sources might we use to understand how reservoirs and natural lakes vary in size in these three states? 
## Part two
### Subsets 
#### Columns nutr to only keep key info that we want
#### Keep sites with at least 200 observations 
#### Join water quality data to spatial data
### Mean Chlorophyll A map
### What is the correlation between Secchi Disk Depth and Chlorophyll a for sites with at least 200 observations?
#### Why might this be the case? 
### What states have the most data? 
#### Make a lagos spatial dataset that has the total number of counts per site.
#### Join this point dataset to the us_boundaries data. 
#### Group by state and sum all the observations in that state and arrange that data from most to least total observations per state. 
### Is there a spatial pattern in Secchi disk depth for lakes with at least 200 observations?

<!--chapter:end:06-Spatial-Analysis.Rmd-->

# Linear Regressions, Quadratic Fits, Residuals, and Spatial

>"How many data scientists does it take to change a light bulb? That depends. It is really a matter of power."

This assignment combined several methods to look at relationships between crop yields and weather data over time.

Data from [USDA National Agricultural Statistical Service (NASS)](https://quickstats.nass.usda.gov/). Assignment by Dr. Matthew Ross and Dr. Nathan Mueller of Colorado State University.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(R.matlab) #pull in matlab files
library(rnassqs) #access to USDA national agriculture data
library(RColorBrewer)
```

## Weather Data Analysis

### Load the PRISM daily maximum temperatures

```{r tmax data}

# daily max temperature
# dimensions: counties x days x years
prism <- readMat("prismiowa.mat")

# look at county #1
t_1981_c1 <- prism$tmaxdaily.iowa[1,,1] #first county, all days, first year
t_1981_c1[366] #check for leap year (366 days)
plot(1:366, t_1981_c1, type = "l") #base r plot

```

```{r tidying up}

# assign dimension names to tmax matrix
dimnames(prism$tmaxdaily.iowa) <- list(prism$COUNTYFP, 1:366, prism$years) #add dimension names

# converted 3d matrix into a data frame
tmaxdf <- as.data.frame.table(prism$tmaxdaily.iowa)

# relabel the columns
colnames(tmaxdf) <- c("countyfp","doy","year","tmax") #name columns
tmaxdf <- tibble(tmaxdf) #tidyverse table

```

### Download NASS corn yield data

```{r yield download, message=FALSE}

# set our API key with NASS
nassqs_auth(key = "B9113AF8-85C4-3CEE-8D93-6E885D49E24F") #Here put in API code from USDA QuickStats service

# parameters to query on 
params <- list(commodity_desc = "CORN", util_practice_desc = "GRAIN", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
cornyieldsall <- nassqs_yields(params)

cornyieldsall$county_ansi <- as.numeric(cornyieldsall$county_ansi)
cornyieldsall$yield <- as.numeric(cornyieldsall$Value)

# clean and filter this dataset
cornyields <- select(cornyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
cornyields <- tibble(cornyields)

```

## Extract Winneshiek County corn yields, fit a linear time trend, make a plot. Is there a significant time trend?
```{r, warning=FALSE, fig.cap="Linear regression of corn yields over time (years) in Winneshieck County, Iowa."}

winnecorn <- cornyields %>%
  filter(county_ansi == "191")

cornlm <- lm(yield ~ year, data = winnecorn)
summary(cornlm) #P=1.77e-13 R^2= 0.755

ggplot(winnecorn, mapping = aes(x = year, y = yield)) +
  geom_point() +
  theme_bw() +
  labs(x = "Year", y = "Corn Yield") +
  geom_smooth(method = lm, se=TRUE, color="#78917E", fill="#C5DDB3")

```

There is a significant positive correlation between corn yields and years in Winneshieck County, with an R-squared value of 0.755 and a P-value of 1.77e-13.


## Fit a quadratic time trend (i.e., year + year^2) and make a plot. Is there evidence for slowing yield growth? 
```{r, warning=FALSE, fig.cap="Quadratic fit of corn yields over time (years) in Winneshieck County, Iowa."}

winnecorn$yearsq <- winnecorn$year^2 #square explanatory variables for quadratic

lm_cornquad <- lm(yield ~ year + yearsq, winnecorn)
summary(lm_cornquad)
winnecorn$y_fitted <- lm_cornquad$fitted.values

#with the fitted values, create a non-linear trend
ggplot(winnecorn) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = y_fitted)) +
  theme_bw() +
  labs(x = "Year", y = "Corn Yield")

```

When we fit a quadratic line to the data, we find that it follows very closely to a linear regression, suggesting a fairly linear relationship between corn yields and years in Winneshieck County. There is no evidence of slowing yield growth in the model. 

## Time Series: Let's analyze the relationship between temperature and yields for the Winneshiek County time series. Use data on yield and summer avg Tmax. Is adding year or Tmax^2 to your model helpful? Make a plot and interpret the results.
```{r, warning=FALSE, fig.cap="Comparative quadratic fit of corn yields over time (blue) and fitted line with maximum summer temperatures as well (red) in Winneshieck County, Iowa."}

# Winneshiek County summer temp maxes
tmaxdf$doy <- as.numeric(tmaxdf$doy) 
tmaxdf$year <- as.numeric(as.character(tmaxdf$year))
tmaxdf$tmax <- as.numeric(tmaxdf$tmax)

winnesummer <- tmaxdf %>%
  filter(countyfp==191 & doy >= 152 & doy <= 243) %>% #day 152= June 1, 243= Aug 31
  group_by(year) %>%
  summarize(meantmax = mean(tmax))

lm_summertmax <- lm(meantmax ~ year, winnesummer)
summary(lm_summertmax) #not sig

winnesummer$yearsq <- winnesummer$year^2 #square explanatory variables for quadratic
winnesummer$tmaxsq <- winnesummer$meantmax^2 

lm_summerquad <- lm(meantmax ~ year + yearsq, winnesummer)
summary(lm_summerquad)
winnesummer$t_fitted <- lm_summerquad$fitted.values

# Join yield and temp data
winne <- inner_join(winnecorn, winnesummer)

lmwinne <- lm(yield ~ yearsq + tmaxsq, data = winne)
summary(lmwinne)
winne$allfit <- lmwinne$fitted.values

ggplot(winne) +
  geom_point(mapping = aes(x = year, y = yield)) +
  geom_line(mapping = aes(x = year, y = allfit, color="red")) +
  geom_line(mapping = aes(x = year, y = y_fitted, color="blue")) +
  theme_bw() +
  scale_colour_manual(name = "Model", 
         values =c("red"="red","blue"="blue"), labels = c("Fit with Max Temp and Year", "Quadratic Yield Fit")) +
  labs(x = "Year", y = "Corn Yield")

```

Adding maximum temperature trends to the model shows a similar trend, but peaks and dips in the fitted line highlight some of the outlying yield values and suggest an underlying relationship between maximum temperatures and yields. However, the relationship between squared maximum temperature and yield has a P-value of 0.14, compared with the year squared P-value of 9.22e-12, so it is clearly not the important driver of trends. This model has an R-squared value of 0.749, around the same as (slightly lower than) the simple linear regression model with only yield vs. year. Thus, adding temperature doesn't significantly add to our understanding of yield trends in Winneshieck County.


## Cross-Section: Analyze the relationship between temperature and yield across all counties in 2018. Is there a relationship? Interpret the results.
```{r, warning=FALSE, fig.cap="Quadratic fit of corn yields versus maximum summer temperatures (Degrees C) across Iowa."}

corn2018 <- cornyields %>%
  filter(year == "2018") %>%
  mutate_at(vars(county_ansi), funs(factor))
  
tmax2018 <- tmaxdf %>%
  filter(year == "2018") %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(countyfp) %>%
  rename("county_ansi" = "countyfp") %>%
  summarize(meantmax = mean(tmax))

yieldtemp_2018 <- inner_join(corn2018, tmax2018, by="county_ansi") %>%
  mutate(tmaxsq = (meantmax^2))

yt_lm <- lm(yield ~ meantmax + tmaxsq, data = yieldtemp_2018)
summary(yt_lm)
yieldtemp_2018$ytfit <- yt_lm$fitted.values

ggplot(yieldtemp_2018) +
  geom_point(mapping = aes(x = meantmax, y = yield)) +
  geom_line(mapping = aes(x = meantmax, y = ytfit, color="red")) +
  theme_bw() + theme(legend.position="none") +
  labs(x = "Mean Max Temperature (C)", y = "Corn Yield")

```

There is a clear relationship with maximum temperatures and corn yields demonstrated in Figure 4. As we might expect, there appears to be a "sweet spot" in regard to temperature, with corn crops performing best at moderate temperatures and yields falling off at both low and high temperature years. Lower mean maximum temperatures may indicate even lower temperatures that can shock crops, and high means are likely to cause high evaporation and withering. P<0.003 for the relationship between temperature and corn yield across Iowa.


## Panel: One way to leverage multiple time series is to group all data into what is called a "panel" regression. Convert the county ID code ("countyfp" or "county_ansi") into factor using as.factor, then include this variable in a regression using all counties' yield and summer temperature data. How does the significance of your temperature coefficients (Tmax, Tmax^2) change? Make a plot comparing actual and fitted yields and interpret the results of your model.
```{r, warning=FALSE, fig.cap="Fitted model yield values versus actual yield values for all counties of Iowa over all available years, from 1981 to 2018."}

corn_all <- cornyields %>%
  mutate_at(vars(county_ansi), funs(factor))
  
tmax_all <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243) %>%
  group_by(countyfp, year) %>%
  rename("county_ansi" = "countyfp") %>%
  summarize(meantmax = mean(tmax))

yieldtemp_all <- inner_join(corn_all, tmax_all) %>%
  mutate(tmaxsq = (meantmax^2)) %>%
  mutate(yearsq = year^2)

ytc_lm <- lm(yield ~ year + meantmax + tmaxsq + county_ansi, data = yieldtemp_all)
summary(ytc_lm)
yieldtemp_all$fittedyield <- ytc_lm$fitted.values

ggplot(yieldtemp_all, mapping = aes(x = fittedyield, y = yield)) +
  geom_point() +
  geom_smooth(method="lm") +
  theme_bw() + theme(legend.position="none") +
  labs(x = "Fitted Yield", y = "Actual Yield")

```

```{r, warning=FALSE, fig.cap="Residuals (top left), Normal Q-Q (top right), Scale-Location (bottom left), and Cook's Distance (bottom right) plots for the panel regression model."}

par(mfrow=c(2,2))
plot(ytc_lm)

```

As a panel regression of all counties over all years, the statistical significance of year, mean maximum temperature, and squared maximum temperature as predictors of yield becomes stronger (P<2e-16 for each). The R squared value for the model is 0.721, indicating a pretty good fit, as is evident in Figure 5. However, the residuals for the model are pretty wide (Figures 5, 6) and the data may not be very normally distributed (Figure 6).


## Soybeans: Download NASS data on soybean yields and explore either a time series relationship for a given county, the cross-sectional relationship for a given year, or a panel across all counties and years.
```{r, warning=FALSE, fig.cap="Soy yields over time (years) across all counties of Iowa, with a panel fit (orange), and linear fit (blue). Panel regression R squared is 0.660, with p<8.5e-06 for year and years squared versus yield."}

# parameters to query on 
params2 <- list(commodity_desc = "SOYBEANS", prodn_practice_desc = "ALL PRODUCTION PRACTICES", year__GE = 1981, state_alpha = "IA")

# download
soyyieldsall <- nassqs_yields(params)

soyyieldsall$county_ansi <- as.numeric(soyyieldsall$county_ansi)
soyyieldsall$yield <- as.numeric(soyyieldsall$Value)

# clean and filter this dataset
soy <- select(soyyieldsall, county_ansi, county_name, yield, year) %>%
  filter(!is.na(county_ansi) & !is.na(yield))
soy <- tibble(soy)

soy_panel <- soy %>%
  mutate_at(vars(county_ansi), funs(factor)) %>%
  mutate(yearsq = year^2)

soypanel_lm <- lm(yield ~ year + yearsq + county_ansi, data = soy_panel)
summary(soypanel_lm)
soy_panel$fittedyield <- soypanel_lm$fitted.values

ggplot(soy_panel, mapping = aes(x = year, y = yield)) +
  geom_point() +
  geom_line(mapping = aes(x = year, y = fittedyield, color="red")) +
  geom_smooth(method="lm") +
  theme_bw() + theme(legend.position="none") +
  labs(x = "Year", y = "Soy Yield")

```

Like with corn yields, soy yields in Iowa follow an upward trend over time, though with wide residuals (Figure 7). 

## Bonus: Find a package to make a county map of Iowa displaying some sort of information about yields or weather. Interpret your map.
```{r, warning=FALSE}
library(sf) #Spatial package that can read and create shapefiles 
library(mapview) #Interactive maps
library(USAboundaries) #USA states and counties

counties <- us_counties

Iowa_ct <- counties(states = 'iowa')
#str(Iowa_ct)
#mapview(Iowa_ct)

summer2018 <- tmaxdf %>%
  filter(doy >= 152 & doy <= 243, year==2018) %>% #day 152= June 1, 243= Aug 31
  group_by(countyfp, year) %>%
  summarize(meantmax = mean(tmax))

Itemp_2018 <- merge(Iowa_ct, summer2018)

mapview(Itemp_2018, zcol ='meantmax', col.regions=brewer.pal(9, "OrRd"), layer.name = "Mean Max. Summer Temp. (C)")

```

<!--chapter:end:07-Linear-Quadratic.Rmd-->

